Operational Feature Selection with Gaussian Mixture Models for remote sensing images

### 2016-10-03 paper.tex
update

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Operational Feature Selection in Gaussian Mixture Models}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Adrien~Lagrange,~Mathieu~Fauvel~and~Manuel~Grizonnet% <-this % stops a space
\thanks{A. Lagrange and M. Fauvel are with the Université de Toulouse,
INP-ENSAT, UMR 1201 DYNAFOR, France and with the INRA, UMR 1201
DYNAFOR, France.}
\thanks{M. Grizonnet is with Centre Nat. d'Etudes Spatiales, French Space Agency, Toulouse, France.}}

@@

Finally, the method is tested and also compared to other classifiers using two different datasets, one of hyperspectral images with a lot of spectral variables and one with heterogeneous spatial features. The results validate the fact that the method performs well in terms of processing time and classification accuracy in comparison to the standard classifiers available in OTB.\\
\end{abstract}

@@

However, processing this data is more and more challenging because of statistical and computational issues. This statistical issues are often referred as the \emph{curse of dimensionality}. The Hughes phenomenon \cite{hughes1968mean} states that with a given number of samples, prediction accuracy will decays when the complexity is higher than some optimum value. The problem is the fast increase of the number of parameters to estimate in order to build a model when dimension expands \cite{bouveyron2014model}. Thus, a important number of labeled samples is needed to perform learning. For example, in the case of Gaussian Mixture Models, the number of parameters progresses quadratically with the dimension, e.g. with 200 describing variables, the model estimation requires at least around 20,000 samples by class.

The computational issues are multiple. The computing infrastructure needed to process data is more and more expensive because the processing might requires a GPU and a large amount of RAM to load images which can weight a dozen of gigabytes \cite{christophe2011remote}\cite{plaza2011high}. The processing time is also limiting and requires to use parallelize computing (GPU, multi-threading).

A possible method to solve these issues is to perform a reduction of dimension. It is possible to extract a set of relevant features to get a parsimonious representation of the data \cite{jimenez1998supervised}. For instance, in land-cover classification, given a set of spatial, temporal and spectral features, it is possible to extract those which are the most discriminant for the purpose of classification \cite{fassnacht2014comparison}. In hyperspectral data analysis from the hundreds of available spectral channels, it is possible to reduce the number of channels to make the processing more efficient in terms of statistical complexity because of the reduction of the number of parameters to estimate and thus also in term of computational time.

@@

There are two ways to reduce dimension \cite{Guyon:2006:FEF:1208773}: features extraction and feature selection. Feature extraction means reformulate and summarize the information by creating new features in combining the existing ones. Linear combinations of the initial features can be extracted as in Principal Component Analysis \cite{jimenez1998supervised} or in independent component analysis \cite{villa2011hyperspectral}. Supervised extraction method has also been investigate such as Fischer discriminant analysis. To the contrary, features selection selects a subset of existing features identified as the most relevant for a given criterion. It has the advantage to be much more interpretable for the end-user. The selected subset of features corresponds to the most important features for the given task.

@@

The second sort are known as \emph{wrapper methods} which can be seen as search method to determine the best subset of variables for a given learning model. As exhaustive searches are too expensive in terms of processing time, several search strategies have been designed. Some are optimal under particular hypothesis as the branch and bound algorithm \cite{narendra1977branch} optimal under the condition of a monotonic criterion function and other sub-optimal but easier to set up. Whitney \cite{whitney1971direct} proposed a forward search adding feature one by one \cite{whitney1971direct} when Marill and Green \cite{marill1963effectiveness} use a backward search eliminating features one by one. Some papers propose to combine these two search method \emph{somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model but on the other hand, as they require the training of multiple models to test various set of variables, they tend to be more time consuming. The proposed in this paper is actually a wrapper method which has been optimized in order to accelerate model learning and testing which is the weak point of this family.

@@

However, this large diversity of methods for feature selection usually do not scale well with the number of pixels to be processed \cite{fauvel2015fast}. Nevertheless, methods based on Gaussian Mixture Models (GMM) have several interesting properties that make them suitable for feature selection in the context of large amount of data. By taking advantage of their intrinsic properties, it is possible to increase the computational efficiency with respect to standard implementation.

Several strategy have been explored to perform feature selection and, among them, wrapper methods receive a certain interest of the community. Wrapper methods can be seen as a search method to determine the best subset of variables for a given learning model. As exhaustive searches are out of question in a practical amount of time, numerous search strategies have been designed. Some optimal are under particular hypothesis \cite{narendra1977branch} and other suboptimal but easier to set up \cite{whitney1971direct}\cite{somol1999adaptive}. Such methods have the advantage to tune the selection to make the most of a particular classifier but require the training of multiple models to test various set of variables which can make them slow if the training is not optimized.

@@

The remaining of this article is organized as follows. Section~\ref{sec:gmm-hd} presents Gaussian Mixture Model classifiers and the features selection methods used to make them suitable for high-dimension space. The work done to develop a smart implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. And finally, the tests conducted to explore variations of the algorithm and to compare it to other standard classifiers are detailed in Section~\ref{sec:test}.

@@

    It  is  referred  to  as  Gaussian  mixture  model  (GMM).   In  a
    supervised    learning    framework,    the    class    parameters
    $\boldsymbol{\mu}_c$,   $\boldsymbol{\Sigma}_c$   and  the   prior
    $\pi_c$ are  usually estimated  through the  conventional unbiased
    empirical estimators:
    \begin{align}
        \hat{\pi}_c &= \frac{n_c}{n},\\
        \hat{\boldsymbol{\mu}}_c &= \frac{1}{n_c} \sum_{\{i|y_i = c\}} \mathbf{x}_i ,\\
        \hat{\boldsymbol{\Sigma}}_c &= \frac{1}{(n_c - 1)} \sum_{\{i|y_i = c\}} (\mathbf{x}_i - \boldsymbol{\mu}_c) (\boldsymbol{}x_i - \boldsymbol{\mu}_c)^t.
    \end{align}

    To predict the  class of a new unseen sample,  the maximum \emph{a
      posteriori}  rule  is  used:
    \begin{equation*}
        \mathbf{x} \text{ belongs to } c \Leftrightarrow c = \text{arg} \max_{c \in C} p(c) p(\mathbf{x}|c).
    \end{equation*}
    Under the GMM,  and identifying $p(c)$ as $\pi_c$  and $p(\mathbf{x}|c)$ as
    $f_c(\mathbf{x}|\theta)$ and by taking the log, a decision rule is obtained
    \begin{eqnarray}\label{eq:decision}
      Q_c(\mathbf{x}) &=& 2 \log \left( p(c) p(\mathbf{x}|c) \right) \nonumber \\
                      &=& - (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \nonumber \\
                      & &-\log (|\boldsymbol{\Sigma}_c|) + 2 \log (\pi_c) - d \log (2\pi).
    \end{eqnarray}

@@

    The computation of~\ref{eq:decision} requires the inversion of the
    covariance matrix and and the  computation of the logarithm of the
    determinant.  The estimations of these terms suffer from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of  samples
    $n_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the cost to collect ground-truth.

@@

        The classification rate is  estimated by a $k$- cross-validation ($k$-CV)~\cite{opac-b1127878}.  To compute  the $k$-CV, a subset is removed  from $\mathcal{S}$ and  the GMM is learned  with the remaining training samples.   A test error is  computed with the removed  training  samples  used  as  validation  samples.   The process is  iterated $k$ times and  the estimated classification rate is computed as the mean  test error over the $k$ subsets of $\mathcal{S}$.

@@

        \subsubsection{Similarity between distributions}
        The similarity  between two distributions is  quantified using
        divergence  measure~\cite{opac-b1097517}. Contrary  to measure
        of correct  classification, divergences are  computed directly
        on the trained model, with no need of validation samples.  Two
        particular   divergences   are   used  in   this   work:   the
        \emph{Kullback–Leibler}        divergence       and        the
        \emph{Jeffries–Matusita}  distance.  The  advantage  of  these
        divergences is  that they have  an explicit expression  in the
        case of Gaussian models. The  simplification allows to get rid
        of  any integration  calculus which  is a  major problem  when
        dealing with high-dimensional data.

        \emph{The   Kullback-Leibler   divergence}   (KL   divergence)
        measures  the  amount  of  information  lost  when  the  first
        distribution     is     approximated     by     the     second
        one\cite{kullback1987letter}. It can be explicitly computed in
        the case of Gaussian distributions:
        \begin{align}
            &KL_{ij} = \frac{1}{2} \biggl( \text{Tr} (\boldsymbol{\Sigma}_{c_i}^{-1} \boldsymbol{\Sigma}_{c_j}) \nonumber \\
            & + (\boldsymbol{\mu}_{c_i} - \boldsymbol{\mu}_{c_j})^t \boldsymbol{\Sigma}_{c_i}^{-1} (\boldsymbol{\mu}_{c_i} - \boldsymbol{\mu}_{c_j}) - d + \log \left( \frac{|\boldsymbol{\Sigma}_{c_i}|}{|\boldsymbol{\Sigma}_{c_j}|} \right) \biggr),
        \end{align}
        where Tr is the trace operator and $d$ the dimension of the distribution.

        The    KL   divergence    is   not    symmetric,   \emph{i.e.}
        $KL_{ij} \ne KL_{i,j}$.  A the symmetrized version  is used to
        compute the criterion function:
        \begin{align}
          SKL_{ij} &=KL_{ij} + KL_{i,j} \nonumber \\
            &= \frac{1}{2} \biggl( \text{Tr} (\boldsymbol{\Sigma}_{c_i}^{-1} \boldsymbol{\Sigma}_{c_j} + \boldsymbol{\Sigma}_{c_j}^{-1} \boldsymbol{\Sigma}_{c_i}) \nonumber \\
            &~~+ (\boldsymbol{\mu}_{c_i} - \boldsymbol{\mu}_{c_j})^t (\boldsymbol{\Sigma}_{c_i}^{-1} + \boldsymbol{\Sigma}_{c_j}^{-1}) (\boldsymbol{\mu}_{c_i} - \boldsymbol{\mu}_{c_j}) - 2d \biggr).
        \end{align}

        The extension  to the multiclass  situation is done  by taking
        the weighted mean  of the KL divergences computed  on all pair
        of classes~\cite{bruzzone1995extension}:
        \begin{equation}
            C_{SKL} = \sum_{i=1}^{C} \sum_{j=i + 1}^{C} \pi_{c_i} \pi_{c_j} SKL_{ij}.
        \end{equation}

@@

        Similar to the KL divergence,  a weighted mean of the distance
        between two classes is computed to aggregate the measures in a
        single value:
        \begin{equation}
            {C}_{JM} = \sum_{i=1}^{C} \sum_{j=i + 1}^{C} \pi_{c_i} \pi_{c_j} {JM}_{ij}.
        \end{equation}

@@

        \begin{table}[!t]
            \centering
            \caption{Summary of the different criterion functions.\label{tab:crit}}
            \begin{tabular}[b]{lccc}
              \toprule
              Criterion & Type Divergence & Complexity \\
              \midrule
              Overall accuracy            & Accuracy   & High \\
              Cohen's kappa               & Accuracy   & High\\
              F1 mean                     & Accuracy   & High\\
              \midrule
              Kullback-Leibler divergence & Divergence & Low \\
              Jeffries-Matusita distance  & Divergence & Low \\
              \bottomrule
            \end{tabular}
        \end{table}

        \subsection{Selection method}
        \label{sec:selection:method}
        Two  sequential  search  functions were  implemented  in  this
        work~\cite{Guyon:2006:FEF:1208773}:  the   sequential  forward
        selection and the sequential  floating forward.  The later one
        is  an   extension  of   the  former.  Both   select  features
        iteratively.

@@

        The Sequential Forward Selection (SFS) starts with an empty set of selected features.  At each  step, the feature that most  improves one selected criterion function $J$ selected in Table~\ref{tab:crit} is added  to the set. This feature is definitively added to the pool of selected features and the algorithm stops when a given number of variables \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sfs} presents the process in details.

@@

        The SFFS works like the SFS but between each step of the SFS algorithm, a backward selection is operated to identify the less important feature. If the criterion function is higher than the best value value ever obtained with a set of same size, the identified feature is picked out. The SBS step is repeated while removing the less important feature leads to an increase of the criterion function. Then SFS is going back. The algorithm stops when a given number of features \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sffs} provides details of the process.

@@

    \subsection{Statistical update rules}
    In the presented method, the most costly operation is the evaluation of the criterion function which needs to be done for each possible subset of variables augmented by one of the candidate variable. To compute these values, the most costly operation are the inversion of the covariance matrix and the computation of its determinant. Both of these operations have a $O(n^3)$ computational complexity. Additionaly, the estimation of classification rate with cross-validation is a heavy process which take a consequent amount of time.

    A major contribution of this work is the optimization in term of computational efficiency. More precisely, three steps of the proposed method has been upgraded in order to reduce computational time. First of all, the GMM model is learned only once using samples. When a covariance matrix or a mean vector of a reduced set of variables is required, it is obtained from the global model learned at the beginning by marginalization.

    Secondly, when a cross-validation is performed, submodels, i.e. GMM model trained with $(n_{cv}-1)$ folds, are not learned using the $(n_{cv}-1)$ folds but, instead, are derived from the global model and from the covariance matrices and mean vectors of the fold used for validation for the given submodel. The process is described in details in Section~\ref{sec:update-cv}.

    Finally and most importantly, when variables are tested one by one during a selection step, costly operations are made when computing criterion functions for each variable especially the computation of the inverse of covariance matrices and its determinant. Several update rules are set up which allow to compute this inverse and determinant only once to test all variables. These update rules are presented in Section~\ref{sec:update-crit}.

@@

        Based on \cite{fauvel2015fast}, a method to accelerate the $n_{cv}$-fold cross-validation process in the case of criterion functions based on correct classification measures was implemented. The idea is to estimate the GMM model with the whole training set once and then, instead of training models on $(n_{cv}-1)$ folds, parameters of the complete model are used to derive those of submodels.

        The following formulae can be obtained
        \begin{prop}
            \label{eq:update-cv1}
            (Cross-validation mean update)
            \begin{equation*}
                \boldsymbol{\mu}_c^{n_c-\nu_c} = \frac{n_c \boldsymbol{\mu}_c^{n_c} - \nu_c \boldsymbol{\mu}_c^{\nu_c}}{n_c - \nu_c} \nonumber
            \end{equation*}
        \end{prop}
        \begin{prop}
            \label{eq:update-cv2}
            (Cross-validation covariance matrix update)
            \begin{align*}
                \boldsymbol{\Sigma}_c^{n_c-\nu_c} = &\frac{1}{n_c-\nu_c-1} \biggl( (n_c-1) \boldsymbol{\Sigma}_c^{n_c} - (\nu_c-1) \boldsymbol{\Sigma}_c^{\nu_c} \nonumber \\
                &- \frac{n_c \nu_c}{(n_c-\nu_c)} (\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})(\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})^t \biggr) \nonumber
            \end{align*}
        \end{prop}
        where $n_c$ is the number of samples of class $c$, $\nu_c$ is the number of samples of class $c$ removed from the initial set, exposants on $\boldsymbol{\Sigma}_c$ and $\boldsymbol{\nu}_c$ denotes the set of samples used to compute them.

@@

    The developed method has been made available as a remote module of Orfeo toolbox called \emph{otbFFSforGMM} and can be download on Github\footnote{\url{https://github.com/Laadr/otbExternalFastFeaturesSelection}}.

    The Orfeo Toolbox is an open-source library for remote sensing image processing. This library allows developers to code external modules and make them available to the community. It is to note that the toolbox already implements several classifiers inherited mostly from OpenCV. Nevertheless, the GMM classifier currently in place is not robust to high dimensionality and that is why a new GMM model has been developed.

    In addition of the C++ class directly available to develop C++ code, three command line application has been produced. These applications allow the user to use and parametrize the algorithm as desire.

    The first application \emph{otbcli\_TrainGMMApp} trains a GMM classifier from multiple pairs of images and training vector data and creates a model file. It also allows to perform ridge regularization with an embedded selection of the regularization parameter among a list of proposed values. The regularization values is chosen by maximizing a criterion function (Cohen's kappa, overall accuracy or mean of F1 scores) estimated by cross-validation.

>>>>

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Operational Feature Selection with Gaussian Mixture Models for remote sensing images}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Adrien~Lagrange,~Mathieu~Fauvel~and~Manuel~Grizonnet% <-this % stops a space
\thanks{A. Lagrange and M. Fauvel are with the Universit\'{e} de Toulouse,
INP-ENSAT, UMR 1201 DYNAFOR, France and with the INRA, UMR 1201
DYNAFOR, France.}
\thanks{M. Grizonnet is with Centre National d'\'{E}tudes Spatiales, French Space Agency, Toulouse, France.}}

@@

Finally, the method is compared to standard classifiers using two different datasets, one of hyperspectral images with a lot of spectral variables and one with heterogeneous spatial features. The results validate the fact that the method performs well in terms of processing time and classification accuracy in comparison to the standard classifiers available in OTB.\\
\end{abstract}

@@

However, processing such data is more and more challenging because of statistical and computational issues. This statistical issues are often referred as the \emph{curse of dimensionality}. The Hughes phenomenon \cite{hughes1968mean} states that with a given number of samples, prediction accuracy will decays when the complexity, \emph{i.e.} the dimension, is higher than some optimum value. The problem is the fast increase of the number of parameters to estimate in order to build a model when dimension expands \cite{bouveyron2014model}. Thus, an important number of labeled samples is needed to perform learning. For example, in the case of Gaussian Mixture Models, the number of parameters progresses quadratically with the dimension, e.g. with 200 describing variables, the model estimation requires at least around 20,000 samples by class.

The computational issues are also multiple. The computing infrastructure needed to process data is more and more expensive because the processing might requires a GPU and a large amount of RAM to load images which can weight a dozen of gigabytes \cite{christophe2011remote}\cite{plaza2011high}. The processing time is also limiting and requires to use parallelize computing (GPU, multi-threading).

A possible method to solve these issues is to perform a reduction of dimension. It is possible to extract a set of relevant features to get a parsimonious representation of the data \cite{jimenez1998supervised}. For instance, in land-cover classification, given a set of spatial, temporal and spectral features, it is possible to extract those which are the most discriminant for the purpose of classification \cite{fassnacht2014comparison}. In hyperspectral data, from the hundreds of available spectral channels, it is possible to reduce the number of channels to make the processing more efficient in terms of statistical complexity because of the reduction of the number of parameters to estimate and thus also in term of computational time.

@@

There are two ways to reduce dimension \cite{Guyon:2006:FEF:1208773}: features extraction and feature selection. Feature extraction means reformulate and summarize the information by creating new features in combining the existing ones. Linear combinations of the initial features can be extracted using for example in Principal Component Analysis \cite{jimenez1998supervised} or Independent Component Analysis \cite{villa2011hyperspectral}. Supervised extraction method has also been investigate such as Fischer discriminant analysis. To the contrary, features selection selects a subset of existing features identified as the most relevant for a given criterion. It has the advantage to be much more interpretable for the end-user. The selected subset of features corresponds to the most important features for the given task.

@@

The second sort are known as \emph{wrapper methods} which can be seen as search method to determine the best subset of variables for a given learning model. As exhaustive searches are too expensive in terms of processing time, several search strategies have been designed. Some are optimal under particular hypothesis as the branch and bound algorithm \cite{narendra1977branch} optimal under the condition of a monotonic criterion function. Others are sub-optimal but easier to set up. Whitney \cite{whitney1971direct} proposed a forward search adding feature one by one \cite{whitney1971direct} when Marill and Green \cite{marill1963effectiveness} use a backward search eliminating features one by one. Some papers propose to combine these two search method \cite{somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model but on the other hand, as they require the training of multiple models to test various set of variables, they tend to be more time consuming. The method proposed in this paper is actually a wrapper method which has been optimized in order to accelerate model learning and testing which is the weak point of this family.

@@

Despite a large diversity of methods, feature selection algorithms usually do not scale well with the number of pixels to be processed \cite{fauvel2015fast}. Nevertheless, methods based on Gaussian Mixture Models (GMM) have several interesting properties that make them suitable for feature selection in the context of large amount of data. By taking advantage of their intrinsic properties, it is possible to increase the computational efficiency with respect to standard implementation.

% Several strategy have been explored to perform feature selection and, among them, wrapper methods receive a certain interest of the community. Wrapper methods can be seen as a search method to determine the best subset of variables for a given learning model. As exhaustive searches are out of question in a practical amount of time, numerous search strategies have been designed. Some optimal are under particular hypothesis \cite{narendra1977branch} and other suboptimal but easier to set up \cite{whitney1971direct}\cite{somol1999adaptive}. Such methods have the advantage to tune the selection to make the most of a particular classifier but require the training of multiple models to test various set of variables which can make them slow if the training is not optimized.

@@

The remaining of this article is organized as follows. Section~\ref{sec:gmm-hd} presents Gaussian Mixture Model classifiers and the problems link to high-dimensionality. The features selection methods used to make them suitable for high-dimension space is detailed in Section~\ref{sec:selection}. The work done to develop an efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. And finally, after introducing the two datasets in Section~\ref{sec:datasets}, the tests conducted to explore variations of the algorithm and to compare it to other standard classifiers are shown and commented in Section~\ref{sec:test}.

@@

    It  is  referred  to  as  Gaussian  mixture  model  (GMM).   In  a
    supervised    learning    framework,    the    class    parameters
    $\boldsymbol{\mu}_c$,   $\boldsymbol{\Sigma}_c$   and  the   prior
    $\pi_c$ are  usually estimated  through the  conventional unbiased
    empirical estimators:
    \begin{align}
        \hat{\pi}_c &= \frac{n_c}{n},\\
        \hat{\boldsymbol{\mu}}_c &= \frac{1}{n_c} \sum_{\{i|y_i = c\}} \mathbf{x}_i ,\\
        \hat{\boldsymbol{\Sigma}}_c &= \frac{1}{(n_c - 1)} \sum_{\{i|y_i = c\}} (\mathbf{x}_i - \boldsymbol{\mu}_c) (\boldsymbol{}x_i - \boldsymbol{\mu}_c)^t.
    \end{align}
    To predict the  class of a new unseen sample,  the maximum \emph{a
      posteriori}  rule  is  used:
    \begin{equation*}
        \mathbf{x} \text{ belongs to } c \Leftrightarrow c = \text{arg} \max_{c \in C} p(c) p(\mathbf{x}|c).
    \end{equation*}
    Under the GMM,  and identifying $p(c)$ as $\pi_c$  and $p(\mathbf{x}|c)$ as
    $f_c(\mathbf{x}|\theta)$ and by taking the log, a decision rule is obtained
    \begin{eqnarray}\label{eq:decision}
      Q_c(\mathbf{x}) &=& 2 \log \left( p(c) p(\mathbf{x}|c) \right) \nonumber \\
                      &=& - (\mathbf{x} - \boldsymbol{\mu}_c)^t \boldsymbol{\Sigma}_c^{-1} (\mathbf{x} - \boldsymbol{\mu}_c) \nonumber \\
                      & &-\log (|\boldsymbol{\Sigma}_c|) + 2 \log (\pi_c) - d \log (2\pi).
    \end{eqnarray}

@@

    The computation of~\ref{eq:decision} requires the inversion of the
    covariance matrix and the  computation of the logarithm of the
    determinant.  The estimation of these terms suffers from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of parameters
    $\eta_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    a singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the high cost to collect ground-truth.

@@

        The classification rate is  estimated by a $n_{cv}$- cross-validation ($n_{cv}$-CV)~\cite{opac-b1127878}.  To compute the $n_{cv}$-CV, a subset is removed  from $\mathcal{S}$ and  the GMM is learned with the remaining training samples. A test error is computed with the removed  training  samples  used  as  validation  samples.   The process is  iterated $n_{cv}$ times and  the estimated classification rate is computed as the mean  test error over the $n_{cv}$ subsets of $\mathcal{S}$.

@@

        \subsubsection{Similarity between distributions}
        The similarity  between two distributions is  quantified using
        divergence  measure~\cite{opac-b1097517}. Contrary  to measure
        of correct  classification, divergences are  computed directly
        on the trained model, with no need of validation samples.  Two
        particular   divergences   are   used  in   this   work:   the
        \emph{Kullback-Leibler}        divergence       and        the
        \emph{Jeffries-Matusita}  distance.  The  advantage  of  these
        divergences is  that they have  an explicit expression  in the
        case of Gaussian models. The  simplification allows to get rid
        of  any integration  calculus which  is a  major problem  when
        dealing with high-dimensional data.

        \emph{The   Kullback-Leibler   divergence}   (KL   divergence)
        measures  the  amount  of  information  lost  when  the  first
        distribution     is     approximated     by     the     second
        one\cite{kullback1987letter}. It can be explicitly computed in
        the case of Gaussian distributions:
        \begin{align}
            &KL_{ij} = \frac{1}{2} \biggl( \text{Tr} (\boldsymbol{\Sigma}_i^{-1} \boldsymbol{\Sigma}_j) \nonumber \\
            & + (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j)^t \boldsymbol{\Sigma}_i^{-1} (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j) - d + \log \left( \frac{|\boldsymbol{\Sigma}_i|}{|\boldsymbol{\Sigma}_j|} \right) \biggr),
        \end{align}
        where Tr is the trace operator and $d$ the dimension of the distribution.

        The    KL   divergence    is   not    symmetric,   \emph{i.e.}
        $KL_{ij} \ne KL_{ji}$.  A symmetrized version  is used to
        compute the criterion function:
        \begin{align}
          SKL_{ij} &=KL_{ij} + KL_{i,j} \nonumber \\
            &= \frac{1}{2} \biggl( \text{Tr} (\boldsymbol{\Sigma}_i^{-1} \boldsymbol{\Sigma}_j + \boldsymbol{\Sigma}_j^{-1} \boldsymbol{\Sigma}_i) \nonumber \\
            &~~+ (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j)^t (\boldsymbol{\Sigma}_i^{-1} + \boldsymbol{\Sigma}_j^{-1}) (\boldsymbol{\mu}_i - \boldsymbol{\mu}_j) - 2d \biggr).
        \end{align}

        The extension  to the multiclass  situation is done  by taking
        the weighted mean  of the KL divergences computed  on all pair
        of classes~\cite{bruzzone1995extension}:
        \begin{equation}
            C_{SKL} = \sum_{i=1}^{C} \sum_{j=i + 1}^{C} \pi_i \pi_j SKL_{ij}.
        \end{equation}

@@

        Similar to the KL divergence,  a weighted mean of the distance
        between two classes is computed to aggregate the measures in a
        single value:
        \begin{equation}
            {C}_{JM} = \sum_{i=1}^{C} \sum_{j=i + 1}^{C} \pi_i \pi_j {JM}_{ij}.
        \end{equation}

@@

        \begin{table}[!t]
            \centering
            \caption{Summary of the different criterion functions.\label{tab:crit}}
            \begin{tabular}[b]{lcc}
              \toprule
              Criterion & Type & Complexity \\
              \midrule
              Overall accuracy            & Accuracy   & High \\
              Cohen's kappa               & Accuracy   & High\\
              F1 mean                     & Accuracy   & High\\
              \midrule
              Kullback-Leibler divergence & Divergence & Low \\
              Jeffries-Matusita distance  & Divergence & Low \\
              \bottomrule
            \end{tabular}
        \end{table}

        \subsection{Selection method}
        \label{sec:selection:method}
        Two  sequential  search  algorithms were  implemented  in  this
        work~\cite{Guyon:2006:FEF:1208773}:  the   sequential  forward
        selection and the sequential  floating forward.  The later one
        is  an   extension  of   the  former.  Both   select  features
        iteratively.

@@

        The Sequential Forward Selection (SFS) starts with an empty set of selected features.  At each  step, the feature that most  improves the criterion function $J$ selected in Table~\ref{tab:crit} is added  to the set. This feature is definitively added to the pool of selected features and the algorithm stops when a given number of variables \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sfs} presents the process in details.

@@

        The SFFS works like the SFS but between each step of the SFS algorithm, a backward selection is operated to identify the less important feature. If the criterion value is higher than the best value ever obtained with a set of same size, the identified feature is picked out. The SBS step is repeated while removing the less important feature leads to an increase of the criterion value. Then SFS is called again. The algorithm stops when a given number of features \emph{maxVarNb} has been selected. The Algorithm~\ref{alg:sffs} provides details of the process.

@@

    \subsection{Statistical update rules}
    In the presented method, the most time-consuming operation is the evaluation of the criterion function which needs to be done for each possible subset of variables composed of the selected set augmented by one of the candidate variable. To compute these values, the most costly operations are the inversion of the covariance matrix and the computation of its determinant. Both of these operations have a $O(n^3)$ computational complexity. Additionally, the estimation of classification rate with cross-validation is a heavy process which takes a substantial amount of time.

    A major contribution of this work is the optimization in term of computational efficiency. More precisely, three steps of the proposed method have been upgraded in order to reduce computational time. First of all, the GMM model is learned only once using samples. When a covariance matrix or a mean vector of a reduced set of variables is required, it is obtained from the global model learned at the beginning by marginalization.

    Secondly, when a cross-validation is performed, submodels, i.e. GMM trained with $(n_{cv}-1)$ folds, are not learned using the $(n_{cv}-1)$ folds but, instead, are derived from the global model and from the covariance matrices and mean vectors of the fold used for validation for the given submodel. The process is described in details in Section~\ref{sec:update-cv}.

    Finally and most importantly, when variables are tested one by one during a selection step, costly operations are made when computing criterion functions for each candidate set especially the computation of the inverse of covariance matrices and its determinant. Several update rules are set up which allow to compute this inverse and determinant only once to test all sets. These update rules are presented in Section~\ref{sec:update-crit}.

@@

        Based on \cite{fauvel2015fast}, a method to accelerate the $n_{cv}$-fold cross-validation process in the case of criterion functions based on correct classification measures was implemented. The idea is to estimate the GMM with the whole training set once and then, instead of training models on $(n_{cv}-1)$ folds, parameters of the complete model are used to derive those of submodels.

        The following formulae can be obtained
        \begin{prop}
            \label{eq:update-cv1}
            (Cross-validation mean update)
            \begin{equation*}
                \boldsymbol{\mu}_c^{n_c-\nu_c} = \frac{n_c \boldsymbol{\mu}_c^{n_c} - \nu_c \boldsymbol{\mu}_c^{\nu_c}}{n_c - \nu_c} \nonumber
            \end{equation*}
        \end{prop}
        \begin{prop}
            \label{eq:update-cv2}
            (Cross-validation covariance matrix update)
            \begin{align*}
                \boldsymbol{\Sigma}_c^{n_c-\nu_c} = &\frac{1}{n_c-\nu_c-1} \biggl( (n_c-1) \boldsymbol{\Sigma}_c^{n_c} - (\nu_c-1) \boldsymbol{\Sigma}_c^{\nu_c} \nonumber \\
                &- \frac{n_c \nu_c}{(n_c-\nu_c)} (\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})(\boldsymbol{\mu}_c^{\nu_c}-\boldsymbol{\mu}_c^{n_c})^t \biggr) \nonumber
            \end{align*}
        \end{prop}
        \noindent where $n_c$ is the number of samples of class $c$, $\nu_c$ is the number of samples of class $c$ removed from the initial set, exposants on $\boldsymbol{\Sigma}_c$ and $\boldsymbol{\nu}_c$ denotes the set of samples used to compute them.

@@

    The developed method has been made available as a remote module of Orfeo toolbox called \emph{otbFFSforGMM} and can be download on Github\footnote{\url{https://github.com/Laadr/otbFFSforGMM}}.

    The Orfeo Toolbox is an open-source library for remote sensing image processing. This library allows developers to code external modules and make them available to the community. It is to note that the toolbox already implements several classifiers inherited mostly from OpenCV. Nevertheless, the GMM classifier currently in place is not robust to high dimensionality and that is why a new GMM model has been developed in the module.

    In addition of the C++ class directly available to develop C++ code, three commandline applications have been produced. These applications allow the user to use and parametrize the algorithm as desired.

    The first application \emph{otbcli\_TrainGMMApp} trains a GMM classifier from multiple pairs of images and training vector data and creates a model file. It also allows to perform ridge regularization with an embedded selection of the regularization parameter among a list of values proposed by the user. The regularization value is chosen by maximizing a criterion function (Cohen's kappa, overall accuracy or mean of F1 scores) estimated by cross-validation.

    The second application \emph{otbcli\_TrainGMMSelectionApp} also trains a GMM classifier and in addition performs a feature selection algorithm (SFS, SFFS) with various possible criterion (Jeffries-Matusita dist., Kullback-Lieber div., Cohen’s kappa, overall accuracy, mean F1-score).
